### 2020-04-18 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Updated README.md

```{r load-packages, include=FALSE}
# The data package needs to be installed from github repository:
#devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
# Also, ask Fernando if the version of `spatsur` needed for the analysis is available from CRAN or needs special distribution
#library(aemet) # Needs install from github
library(covid19env)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(lubridate)
#library(meteoland) # Ask Fernando if this package was used in the end
library(plm)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

@@

```{r descriptive-statistics, echo=FALSE}
junk <- data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

>>>>

```{r load-packages, include=FALSE}
# The data package needs to be installed from github repository:
# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")

# This version of spsur may be needed:
# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install

library(covid19env)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(lubridate)
#library(meteoland) # Ask Fernando if this package was used in the end
library(plm)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

@@

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

### 2020-04-18 README.Rmd
Updated README.md

---
title: A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature and humidity. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures. The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations. Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.

journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r load-packages, include=FALSE}
# The data package needs to be installed from github repository:
#devtools::install_github("paezha/covid19-environmental-correlates/covid19env")
# Also, ask Fernando if the version of `spatsur` needed for the analysis is available from CRAN or needs special distribution
#library(aemet) # Needs install from github
library(covid19env)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(lubridate)
#library(meteoland) # Ask Fernando if this package was used in the end
library(plm)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

@@

Context, Data, and Methods
============

## Covid-19 in Spain

@@

## Selection of Variables

@@

## Sources of Data

@@

```{r descriptive-statistics, echo=FALSE}
junk <- data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

## Methods: Spatial SUR

@@

## SUR Models

@@

Summary of best model.
```{r summary-best-model, echo=FALSE}
data.frame(Coefficient = sur.slm_lag11$coefficients, 
                   pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
  rownames_to_column(var = "Variable") %>%
  mutate(Variable = c("Intercept", 
                      "log(GDPpc)", 
                      "log(Older)",
                      "log(Density)",
                      "Transit",
                      "log(Humidity)",
                      "log(Temperature)",
                      rep(c("Intercept",
                            "log(Density)", 
                            "Transit", 
                            "log(Humidity)", 
                            "log(Temperature)"), 29)),
         Variable = factor(Variable, 
                           levels = c("Intercept", 
                                      "log(GDPpc)", 
                                      "log(Older)", 
                                      "log(Density)",
                                      "Transit", 
                                      "log(Humidity)",
                                      "log(Temperature)"))) %>%
  group_by(Variable) %>%
  summarize(Min = min(Coefficient), 
            Mean = mean(Coefficient), 
            Max = max(Coefficient),
            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
  mutate(Note = c("Non-constrained", 
                  "Constrained", 
                  "Constrained", 
                  "Non-constrained", 
                  "Non-constrained",
                  "Non-constrained",
                  "Non-constrained")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 3,
        col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
                       "Non-constrained: coefficient was allowed to vary across equations",
                       "Constrained: coefficient as constant across equations"))
```

Discussion
===================

@@

Concluding Remarks
===================

@@

Acknowledgments {#acknowledgments .unnumbered}
==========

Add acknowledgments as appropriate in final draft.

The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `aemet` [], `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].

References {#references .unnumbered}
==========

>>>>

---
output: github_document
always_allow_html: true
bibliography: References.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

Antonio Paez (McMaster University)  
Fernando A. Lopez (Universidad Politecnica de Cartagena)  
Tatiane Menezes (Universidade Federal de Pernambuco)  
Renata Cavalcanti (Universidade Federal de Pernambuco)  
Maira Galdino da Rocha Pitta (Universidade Federal de Pernambuco)

## Abstract

Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature and humidity. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures. The evidence with respect to humidity is more mixed, with coefficients for this variable that are significant in only some equations. Our control variables also yield interesting insights, as population density and percentage of older adults display negative associations with incidence of COVID-19, whereas the presence of mass transit systems in the province is associated with a greater incidence of the disease.

## Keywords

SARS-CoV2  
COVID-19  
Temperature  
Humidity  
Spatial SUR  
Contagion  
Spain

```{r load-packages, include=FALSE}
# The data package needs to be installed from github repository:
# devtools::install_github("paezha/covid19-environmental-correlates/covid19env")

# This version of spsur may be needed:
# Download spsur_1.0.1.3.tar.gz from https://github.com/paezha/covid19-environmental-correlates and install

library(covid19env)
library(ggthemes)
library(gridExtra)
library(kableExtra)
library(lubridate)
#library(meteoland) # Ask Fernando if this package was used in the end
library(plm)
library(sf)
library(spdep)
library(spsur)
library(tidyverse)
library(units)
```

@@

## Context, Data, and Methods

### Covid-19 in Spain

@@

### Selection of Variables

@@

### Sources of Data

@@

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("COVID-19 Incidence",
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                   mean(provinces_spain$GDPpc),
                   mean(provinces_spain$Older),
                   mean(provinces_spain$Density),
                   mean(covid19_spain$Mean_Temp),
                   mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                   sd(provinces_spain$GDPpc),
                   sd(provinces_spain$Older),
                   sd(provinces_spain$Density),
                   sd(covid19_spain$Mean_Temp),
                   sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics") %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

### Methods: Spatial SUR

@@

### SUR Models

@@

Summary of best model.
```{r summary-best-model, echo=FALSE}
data.frame(Coefficient = sur.slm_lag11$coefficients, 
                   pval = dnorm(abs(sur.slm_lag11$coefficients/sur.slm_lag11$rest.se))) %>%
  rownames_to_column(var = "Variable") %>%
  mutate(Variable = c("Intercept", 
                      "log(GDPpc)", 
                      "log(Older)",
                      "log(Density)",
                      "Transit",
                      "log(Humidity)",
                      "log(Temperature)",
                      rep(c("Intercept",
                            "log(Density)", 
                            "Transit", 
                            "log(Humidity)", 
                            "log(Temperature)"), 29)),
         Variable = factor(Variable, 
                           levels = c("Intercept", 
                                      "log(GDPpc)", 
                                      "log(Older)", 
                                      "log(Density)",
                                      "Transit", 
                                      "log(Humidity)",
                                      "log(Temperature)"))) %>%
  group_by(Variable) %>%
  summarize(Min = min(Coefficient), 
            Mean = mean(Coefficient), 
            Max = max(Coefficient),
            pns = sum(pval > 0.10), # Number of coefficients with p > 0.1
            p10 = sum(pval <= 0.10) - sum(pval <= 0.05), # Number of coefficients with  0.1 <= pval < 0.05
            p05 = sum(pval <= 0.05)) %>% # Number of coefficients with pval <=5 
  mutate(Note = c("Non-constrained", 
                  "Constrained", 
                  "Constrained", 
                  "Non-constrained", 
                  "Non-constrained",
                  "Non-constrained",
                  "Non-constrained")) %>%
  kable("html",
        booktabs = TRUE,
        digits = 3,
        col.names = c("Variable", "Min", "Mean", "Max", "p > 0.10", "0.10 <= p < 0.05", "p <= 0.05", "Note"),
        caption = "\\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  add_header_above(c(" " = 1, "Estimates" = 3, "Significance" = 3, " " = 1)) %>%
  footnote(general = c("Significance: This is the number of coefficients with p-values as indicated",
                       "Non-constrained: coefficient was allowed to vary across equations",
                       "Constrained: coefficient as constant across equations"))
```

## Discussion

@@

## Concluding Remarks

@@

## Acknowledgments

The following `R` packages were used in the course of this investigation and the authors wish to acknowledge their developers: `ggthemes` [@Arnold2019], `gridExtra` [@Auguie2017gridextra], `kableExtra` [@Zhu2019], `knitr` [@Xie2014; @Xie2015], `lubridate` [@Grolemund2011dates], `plm` [@Millo2017robust], `rticles` [@Allaire2020], `sf` [@Pebesma2018], `spdep` [@Bivand2013], spsur [@Angulo2020spsur] `tidyverse` [@Wickham2019], `units` [@Pebesma2016].

## References
